---
title: "Exported Data Combination"
output: html_document
---

# Dependencies

**Upstream script:** `Combine_processed_data_w0_Claude.Rmd`

**Input files** (read from `output_dir` = `data_dir/output/`):

- `AMB60-MS_Result_combined.csv`
- `AMB60-A2_Result_combined.csv`

**Output files** (written to `output_dir`):

- `AMB60_{compound_type}_{subfolder}_combined.csv` — merged MS + A2 ambient data with source column
- `AMB60_{compound_type}_{subfolder}_stats.csv` — per-compound statistics broken out by MS, A2, and combined

---

# Setup

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)

# ===== USER PARAMETERS =====
base_dir      <- ""       # Root directory containing ASC files
subfolder     <- ""       # Optional seasonal subfolder (e.g. "Winter2425"); leave "" if files are directly in base_dir

compound_type <- "C6C12"  # "C2C6" or "C6C12"
# ===========================

# Derived paths
data_dir   <- file.path(base_dir, subfolder)
output_dir <- file.path(data_dir, "output")
```

# Functions

```{r functions}
# Calculate summary statistics for a dataset, optionally filtered by source
calculate_stats <- function(data, source_filter = NULL) {
  
  if (!is.null(source_filter)) {
    data <- data %>% filter(source == source_filter)
  }
  
  compound_cols <- setdiff(names(data), c("Sampling date", "source"))
  
  stats <- map_dfr(compound_cols, function(compound) {
    values <- data[[compound]]
    
    tibble(
      compound   = compound,
      n_total    = length(values),
      n_zero     = sum(values == 0, na.rm = TRUE),
      pct_zero   = round(100 * sum(values == 0, na.rm = TRUE) / length(values), 1),
      mean_all   = round(mean(values, na.rm = TRUE), 3),
      median_all = round(median(values, na.rm = TRUE), 3),
      min_all    = round(min(values, na.rm = TRUE), 3),
      max_all    = round(max(values, na.rm = TRUE), 3)
    )
  })
  
  return(stats)
}
```

# Read Input Files

```{r read-files, message=FALSE, warning=FALSE}
ms_file <- file.path(output_dir, paste0("AMB60-MS_Result_combined.csv"))
a2_file <- file.path(output_dir, paste0("AMB60-A2_Result_combined.csv"))

if (!file.exists(ms_file)) stop("MS file not found: ", ms_file)
if (!file.exists(a2_file)) stop("A2 file not found: ", a2_file)

cat("Reading MS file...\n")
data_ms <- read_csv(ms_file, show_col_types = FALSE) %>%
  mutate(source = "MS")

cat("Reading A2 file...\n")
data_a2 <- read_csv(a2_file, show_col_types = FALSE) %>%
  mutate(source = "A2")

cat("MS rows:", nrow(data_ms), "\n")
cat("A2 rows:", nrow(data_a2), "\n")
```

# Combine Datasets

```{r combine, message=FALSE, warning=FALSE}
data_combined <- bind_rows(data_ms, data_a2) %>%
  arrange(`Sampling date`)

cat("Combined rows:", nrow(data_combined), "\n")
```

# Calculate Statistics

```{r statistics, message=FALSE, warning=FALSE}
stats_ms  <- calculate_stats(data_combined, "MS")
stats_a2  <- calculate_stats(data_combined, "A2")
stats_all <- calculate_stats(data_combined)

# Combine into unified report
stats_unified <- stats_ms %>%
  rename_with(~paste0(., "_MS"), -compound) %>%
  left_join(
    stats_a2 %>% rename_with(~paste0(., "_A2"), -compound),
    by = "compound"
  ) %>%
  left_join(
    stats_all %>% rename_with(~paste0(., "_All"), -compound),
    by = "compound"
  ) %>%
  select(compound,
         starts_with("n_total"), starts_with("n_zero"), starts_with("pct_zero"),
         starts_with("mean"), starts_with("median"), starts_with("min"), starts_with("max"))

cat("Statistics calculated for", nrow(stats_unified), "compounds\n")
print(stats_unified, n = Inf)
```

# Export

```{r export, message=FALSE, warning=FALSE}
# Create output directory if it doesn't exist
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)

# Combined data
combined_filename <- paste0("AMB60_", compound_type, "_", subfolder, "_combined.csv")
combined_path     <- file.path(output_dir, combined_filename)
write_csv(data_combined, combined_path)
cat("Exported combined data:", combined_filename, "(", nrow(data_combined), "rows)\n")

# Statistics report
stats_filename <- paste0("AMB60_", compound_type, "_", subfolder, "_stats.csv")
stats_path     <- file.path(output_dir, stats_filename)
write_csv(stats_unified, stats_path)
cat("Exported statistics:", stats_filename, "\n")

cat("\nAll files saved to:", output_dir, "\n")
```
