---
title: "Vistachrom ASC File Combiner"
output: html_document
---

# Dependencies

**Input:** `.Asc` files exported from Vistachrom PeakViewer, located in `data_dir`
Expected filename format: `YYYYMMDD_to_YYYYMMDD_TYPE_MEASURE.Asc`

**Output:** Combined CSV files written to `output_dir` (`data_dir/output/`):

- `AMB60-A2_Result_combined.csv`
- `AMB60-MS_Result_combined.csv`
- `CYL07-A2_RTime_combined.csv`
- `CYL07-MS_RTime_combined.csv`
- `CALIB60M_Result_combined.csv`
- `CALIB60M_RTime_combined.csv`
- `CALIB-MS_Result_combined.csv`
- `CALIB-MS_RTime_combined.csv`
- `ZERO60-2_Result_combined.csv`
- `ZERO-MS_Result_combined.csv`

---

# Setup

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)

# ===== USER PARAMETERS =====
base_dir  <- ""   # Root directory containing ASC files (use / not \)
subfolder <- ""   # Optional seasonal subfolder (e.g. "2025" or "Winter"); leave "" if ASC files are directly in base_dir

compound_type <- "C6C12"  # "C2C6" or "C6C12"
# ===========================

# Derived paths
data_dir   <- file.path(base_dir, subfolder)
output_dir <- file.path(data_dir, "output")
```

# Phase 1: File Discovery & Parsing

```{r file-discovery}
# Get list of all .Asc files in the directory
asc_files <- list.files(data_dir, pattern = "\\.Asc$", full.names = TRUE, ignore.case = TRUE)

cat("Found", length(asc_files), "ASC files\n\n")

# Parse filename to extract metadata
parse_filename <- function(filepath) {
  filename <- basename(filepath)
  
  # Pattern: YYYYMMDD_to_YYYYMMDD_TYPE_MEASURE.Asc
  pattern <- "^(\\d{8})_to_(\\d{8})_([^_]+)_([^.]+)\\.Asc$"
  
  if (str_detect(filename, pattern)) {
    matches <- str_match(filename, pattern)
    
    tibble(
      filepath   = filepath,
      filename   = filename,
      start_date = matches[2],
      end_date   = matches[3],
      type       = matches[4],
      measure    = matches[5],
      date_range = paste(matches[2], "to", matches[3])
    )
  } else {
    # Return NA row if pattern doesn't match
    tibble(
      filepath   = filepath,
      filename   = filename,
      start_date = NA_character_,
      end_date   = NA_character_,
      type       = NA_character_,
      measure    = NA_character_,
      date_range = NA_character_
    )
  }
}

# Create file inventory
file_inventory <- map_dfr(asc_files, parse_filename)

# Display inventory
cat("File Inventory:\n")
print(file_inventory %>% select(filename, type, measure, date_range), n = Inf)

# Check for any files that didn't match the pattern
unmatched <- file_inventory %>% filter(is.na(type))
if (nrow(unmatched) > 0) {
  cat("\n⚠ Warning: The following files did not match the expected naming pattern:\n")
  print(unmatched$filename)
}

# Remove unmatched files from inventory
file_inventory <- file_inventory %>% filter(!is.na(type))

cat("\n", nrow(file_inventory), "files matched the expected pattern\n")
```

# Phase 2: Read & Parse Files

```{r read-files}
# Function to read a single ASC file
# ASC files are tab-delimited with embedded nul characters requiring readLines() + strsplit() parsing
read_asc_file <- function(filepath, source_label) {
  tryCatch({
    # Read raw lines and suppress the nul warnings
    raw_lines <- suppressWarnings(readLines(filepath, warn = FALSE))
    
    if (length(raw_lines) == 0) {
      stop("File is empty or unreadable")
    }
    
    # Parse tab-delimited data manually
    split_lines <- strsplit(raw_lines, "\t")
    
    # Get column names from first line
    col_names <- split_lines[[1]]
    
    if (length(split_lines) > 1) {
      # Convert to matrix then data frame
      data_matrix <- do.call(rbind, split_lines[-1])
      df <- as.data.frame(data_matrix, stringsAsFactors = FALSE)
      colnames(df) <- col_names
      
      # Convert Sampling date to datetime
      df <- df %>%
        mutate(`Sampling date` = as.POSIXct(`Sampling date`,
                                             format = "%Y/%m/%d %H:%M:%S"))
      
      # Convert all other columns to numeric (except Sampling date)
      numeric_cols <- setdiff(names(df), "Sampling date")
      df <- df %>%
        mutate(across(all_of(numeric_cols), as.numeric))
      
      # Add source column
      df <- df %>%
        mutate(source = source_label, .after = `Sampling date`)
      
      return(df)
    } else {
      stop("No data rows found after header")
    }
    
  }, error = function(e) {
    cat("  ✗ Error reading file:", e$message, "\n")
    return(NULL)
  })
}

# Read all files and organize by type and measure
cat("\nReading files...\n")

all_data <- list()

for (i in 1:nrow(file_inventory)) {
  row <- file_inventory[i, ]
  
  cat("Reading:", row$filename, "\n")
  
  df <- read_asc_file(row$filepath, row$date_range)
  
  if (!is.null(df)) {
    type    <- row$type
    measure <- row$measure
    
    if (is.null(all_data[[type]]))           all_data[[type]] <- list()
    if (is.null(all_data[[type]][[measure]])) all_data[[type]][[measure]] <- list()
    
    all_data[[type]][[measure]][[length(all_data[[type]][[measure]]) + 1]] <- df
    
    cat("  ✓ Successfully read:", nrow(df), "rows,", ncol(df), "columns\n")
  }
  
  cat("\n")
}

cat("Finished reading files\n\n")

# Display data structure
cat("Data structure:\n")
for (type in names(all_data)) {
  cat("  ", type, ":\n")
  for (measure in names(all_data[[type]])) {
    n_files <- length(all_data[[type]][[measure]])
    cat("    ", measure, ":", n_files, "file(s)\n")
  }
}
```

# Phase 3: Filter by Interest

```{r filter-data}
# AMB60-A2, AMB60-MS: Result only
# CYL07-A2, CYL07-MS: RTime only
# CALIB60M, CALIB-MS: Result AND RTime
# ZERO60-2, ZERO-MS:  Result only

cat("\nFiltering data by interest...\n\n")

filtered_data <- list()

# AMB types - Result only
for (type in c("AMB60-A2", "AMB60-MS")) {
  if (!is.null(all_data[[type]]) && !is.null(all_data[[type]][["Result"]])) {
    filtered_data[[type]] <- list()
    filtered_data[[type]][["Result"]] <- all_data[[type]][["Result"]]
    cat("✓ Kept", type, "Result:", length(all_data[[type]][["Result"]]), "file(s)\n")
  }
}

# CYL07 types - RTime only
for (type in c("CYL07-A2", "CYL07-MS")) {
  if (!is.null(all_data[[type]]) && !is.null(all_data[[type]][["RTime"]])) {
    filtered_data[[type]] <- list()
    filtered_data[[type]][["RTime"]] <- all_data[[type]][["RTime"]]
    cat("✓ Kept", type, "RTime:", length(all_data[[type]][["RTime"]]), "file(s)\n")
  }
}

# CALIB types - Result AND RTime
for (type in c("CALIB60M", "CALIB-MS")) {
  if (!is.null(all_data[[type]])) {
    filtered_data[[type]] <- list()
    
    if (!is.null(all_data[[type]][["Result"]])) {
      filtered_data[[type]][["Result"]] <- all_data[[type]][["Result"]]
      cat("✓ Kept", type, "Result:", length(all_data[[type]][["Result"]]), "file(s)\n")
    }
    
    if (!is.null(all_data[[type]][["RTime"]])) {
      filtered_data[[type]][["RTime"]] <- all_data[[type]][["RTime"]]
      cat("✓ Kept", type, "RTime:", length(all_data[[type]][["RTime"]]), "file(s)\n")
    }
  }
}

# Zero types - Result only
for (type in c("ZERO-MS", "ZERO60-2")) {
  if (!is.null(all_data[[type]]) && !is.null(all_data[[type]][["Result"]])) {
    filtered_data[[type]] <- list()
    filtered_data[[type]][["Result"]] <- all_data[[type]][["Result"]]
    cat("✓ Kept", type, "Result:", length(all_data[[type]][["Result"]]), "file(s)\n")
  }
}

cat("\nFiltered data structure:\n")
for (type in names(filtered_data)) {
  cat("  ", type, ":\n")
  for (measure in names(filtered_data[[type]])) {
    n_files <- length(filtered_data[[type]][[measure]])
    cat("    ", measure, ":", n_files, "file(s)\n")
  }
}
```

# Phase 4: Combine Files

```{r combine-files}
cat("\nCombining files within each type+measure...\n\n")

combined_data <- list()

for (type in names(filtered_data)) {
  combined_data[[type]] <- list()
  
  for (measure in names(filtered_data[[type]])) {
    df_list <- filtered_data[[type]][[measure]]
    
    if (length(df_list) > 0) {
      combined_df <- bind_rows(df_list)
      combined_data[[type]][[measure]] <- combined_df
      
      cat("✓", type, measure, "- Combined", length(df_list), "file(s) into",
          nrow(combined_df), "total rows\n")
    }
  }
}
```

# Phase 5: Handle Duplicates

```{r handle-duplicates}
cat("\nHandling duplicates...\n\n")

final_data <- list()

# AMB duplicates: average non-zero values; keep non-zero when one value is zero
handle_amb_duplicates <- function(df) {
  compound_cols <- setdiff(names(df), c("Sampling date", "source"))
  
  df_dedup <- df %>%
    group_by(`Sampling date`) %>%
    summarise(
      across(all_of(compound_cols),
             ~{
               vals <- .x[!is.na(.x)]
               if (length(vals) == 0) {
                 NA_real_
               } else if (length(vals) == 1) {
                 vals[1]
               } else {
                 zero_vals <- vals == 0
                 if (all(zero_vals)) {
                   0
                 } else if (any(zero_vals)) {
                   mean(vals[!zero_vals])
                 } else {
                   mean(vals)
                 }
               }
             }),
      .groups = "drop"
    )
  
  return(df_dedup)
}

for (type in names(combined_data)) {
  final_data[[type]] <- list()
  
  for (measure in names(combined_data[[type]])) {
    df <- combined_data[[type]][[measure]]
    
    n_duplicates <- df %>%
      group_by(`Sampling date`) %>%
      filter(n() > 1) %>%
      nrow()
    
    if (str_detect(type, "AMB60")) {
      # AMB: average duplicates and drop source column
      df_final <- handle_amb_duplicates(df)
      cat("✓", type, measure, "- Handled", n_duplicates, "duplicate row(s),",
          "final:", nrow(df_final), "rows (source column dropped)\n")
      
    } else {
      # CYL07 and CALIB: keep all rows with source column
      df_final <- df
      if (n_duplicates > 0) {
        cat("✓", type, measure, "- Kept", n_duplicates, "duplicate row(s),",
            "final:", nrow(df_final), "rows (with source column)\n")
      } else {
        cat("✓", type, measure, "- No duplicates, final:", nrow(df_final), "rows\n")
      }
    }
    
    final_data[[type]][[measure]] <- df_final
  }
}
```

# Phase 6: Export to CSV

```{r export-csv}
cat("\nExporting combined dataframes to CSV...\n\n")

# Create output directory if it doesn't exist
if (!dir.exists(output_dir)) {
  dir.create(output_dir, recursive = TRUE)
}

exported_files <- c()

for (type in names(final_data)) {
  for (measure in names(final_data[[type]])) {
    df <- final_data[[type]][[measure]]
    
    output_filename <- paste0(type, "_", measure, "_combined.csv")
    output_path     <- file.path(output_dir, output_filename)
    
    write_csv(df, output_path)
    
    cat("✓ Exported:", output_filename, "(", nrow(df), "rows)\n")
    exported_files <- c(exported_files, output_filename)
  }
}

cat("\n")
cat(rep("=", 80), "\n", sep = "")
cat("EXPORT COMPLETE\n")
cat(rep("=", 80), "\n", sep = "")
cat("Exported", length(exported_files), "CSV files to:\n")
cat(output_dir, "\n\n")
cat("Files created:\n")
cat(paste("  -", exported_files, collapse = "\n"), "\n")
```

# Summary Statistics

```{r summary-stats}
cat("\n")
cat(rep("=", 80), "\n", sep = "")
cat("SUMMARY STATISTICS\n")
cat(rep("=", 80), "\n\n", sep = "")

for (type in names(final_data)) {
  for (measure in names(final_data[[type]])) {
    df <- final_data[[type]][[measure]]
    
    cat(type, "-", measure, "\n")
    cat("  Rows:", nrow(df), "\n")
    cat("  Columns:", ncol(df), "\n")
    cat("  Date range:", format(min(df$`Sampling date`), "%Y-%m-%d %H:%M"), "to",
        format(max(df$`Sampling date`), "%Y-%m-%d %H:%M"), "\n")
    
    if ("source" %in% names(df)) {
      cat("  Unique sources:", length(unique(df$source)), "\n")
    }
    
    cat("\n")
  }
}
```

# Process Explanation

**Phase 1: File Discovery & Parsing**

- Lists all `.Asc` files in `data_dir`
- Uses regex to extract metadata from filenames (`YYYYMMDD_to_YYYYMMDD_TYPE_MEASURE.Asc`)
- Warns about any files that don't match the expected naming pattern

**Phase 2: Read & Parse Files**

- Reads each file using `readLines()` with nul character suppression
- Manually parses tab-delimited data (splits by `\t`)
- Converts `Sampling date` to POSIXct datetime; all measurement columns to numeric
- Adds a `source` column containing the date range from the filename
- Organizes data in nested list: `all_data[[type]][[measure]]`

**Phase 3: Filter by Interest**

- **AMB60-A2, AMB60-MS**: Result only
- **CYL07-A2, CYL07-MS**: RTime only
- **CALIB60M, CALIB-MS**: Result AND RTime
- **ZERO60-2, ZERO-MS**: Result only

**Phase 4: Combine Files**

- Stacks all dataframes vertically for each type+measure using `bind_rows()`

**Phase 5: Handle Duplicates**

- **AMB files**: Groups by `Sampling date`; averages non-zero values, keeps non-zero when mixed with zero, keeps 0 if all zero. Drops source column.
- **CYL07 and CALIB files**: Keeps all rows as-is with source column intact

**Phase 6: Export to CSV**

- Creates `output/` subdirectory within `data_dir` if it doesn't exist
- Exports each combined dataframe as `[TYPE]_[MEASURE]_combined.csv`
