---
title: "Cylinder ASC Comparison"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

---

## About This Script

This script reads CYL07 ASC files directly from a user-specified folder and produces
a retention time (RTime) comparison report. It evaluates A2 and MS instrument files
separately against a reference compound list, compares A2 vs MS result concentrations,
and checks for data continuity gaps and concentration jumps.

**Required inputs:**

1. A folder containing CYL07 ASC files with names matching the pattern:
   `YYYYMM_CYL07-A2_Result_airmoVOC C2-C6.Asc` (or `C6-C12`, `MS`, etc.)

2. A reference CSV file (`ID_cyl_peaks.csv`) with a column named `Compound`
   listing the expected compounds for the cylinder.

**Output:** `rtime_comparison_report.csv` saved to the specified output folder.

---

## User Configuration

> **START HERE.** Set all parameters in this section before knitting.

```{r user_config}
# ==============================================================================
# USER INPUTS — edit these values only
# ==============================================================================

# Full path to the folder containing CYL07 ASC files
#   Example: "C:/Data/CYL07_files"
cyl_folder <- "PASTE/YOUR/CYL07/FOLDER/PATH/HERE" # note: use / (not \)

# Full path to the reference compound CSV file (ID_cyl_peaks.csv)
#   This file must have a column named "Compound"
#   Example: "C:/Data/ID_cyl_peaks.csv"
reference_file <- "PASTE/YOUR/REFERENCE/FILE/PATH/HERE" # note: use / (not \)

# Full path to the folder where output files will be saved
#   Will be created automatically if it does not exist
output_folder <- cyl_folder # "PASTE/YOUR/OUTPUT/FOLDER/PATH/HERE"  # if you want a different output folder, put that path here

# Compound range being analyzed — must be either "C2C6" or "C6C12"
compound_type <- "C6C12"

# Percent difference threshold for flagging A2 vs MS result comparisons
#   Pairs where abs(A2 - MS) / mean(A2, MS) * 100 exceeds this value are flagged
pct_diff_threshold <- 10

# Gap threshold: flag sampling gaps longer than this many days
gap_threshold_days <- 7

# ==============================================================================
```

---

## Setup

```{r libraries}
library(tidyverse)
library(lubridate)

# ---------- Validate compound_type ----------
if (!compound_type %in% c("C2C6", "C6C12")) {
  stop(
    "compound_type must be 'C2C6' or 'C6C12'. Current value: ", compound_type
  )
}

# ---------- Map compound_type to the suffix used in ASC filenames ----------
# ASC filenames use "C2-C6" or "C6-C12" (with hyphens), not "C2C6" / "C6C12"
asc_suffix <- if (compound_type == "C2C6") "C2-C6" else "C6-C12"

# ---------- Create output folder if needed ----------
if (!dir.exists(output_folder)) {
  dir.create(output_folder, recursive = TRUE)
  cat("Created output folder:", output_folder, "\n")
}

# ---------- Path validation ----------
if (!dir.exists(cyl_folder)) {
  stop("CYL07 folder not found: ", cyl_folder)
}
if (!file.exists(reference_file)) {
  stop("Reference file not found: ", reference_file)
}

cat("CYL folder:       ", cyl_folder, "\n")
cat("Reference file:   ", reference_file, "\n")
cat("Output folder:    ", output_folder, "\n")
cat("Compound type:    ", compound_type, "(ASC suffix:", asc_suffix, ")\n")
cat("% diff threshold:", pct_diff_threshold, "%\n")
cat("Gap threshold:    ", gap_threshold_days, "days\n\n")
```

---

## Helper Functions

```{r helper_read_asc}
# ------------------------------------------------------------------------------
# read_asc_file()
#
# Reads a single ASC file (tab-delimited export from Vistachrom).
# Uses read_delim() with all columns read as character to avoid type coercion
# issues. Removes completely empty rows.
#
# Args:
#   filepath  : Full path to the .Asc file
#
# Returns:
#   A tibble with all columns as character, or NULL if the file does not exist.
# ------------------------------------------------------------------------------

read_asc_file <- function(filepath) {

  if (!file.exists(filepath)) {
    return(NULL)
  }

  # Use readLines with suppressed warnings to handle embedded null characters,
  # which cause read_delim() to fail on some Vistachrom ASC exports.
  raw_lines <- suppressWarnings(readLines(filepath, encoding = "UTF-8"))

  # Remove completely empty lines
  raw_lines <- raw_lines[nzchar(raw_lines)]

  if (length(raw_lines) < 2) {
    warning("File appears empty or has only a header: ", filepath)
    return(NULL)
  }

  # Split each line on tab to get fields
  split_lines <- strsplit(raw_lines, "\t")

  # Use the first line as column names
  col_names <- split_lines[[1]]
  data_rows  <- split_lines[-1]

  # Pad or trim rows to match the number of column names
  n_cols <- length(col_names)
  data_rows <- lapply(data_rows, function(row) {
    length(row) <- n_cols  # shorter rows get NA; longer rows get truncated
    row
  })

  # Build tibble — all columns as character to match read_delim() behavior
  data <- as_tibble(
    do.call(rbind, data_rows),
    .name_repair = "minimal"
  )
  names(data) <- col_names

  # Drop rows where every column is NA
  data <- data %>% filter(!if_all(everything(), is.na))

  return(data)
}
```

```{r helper_parse_filenames}
# ------------------------------------------------------------------------------
# parse_cyl_filenames()
#
# Scans the CYL folder for ASC files matching either of two supported naming
# patterns and extracts instrument and data type from each filename.
#
# Supported patterns:
#   Pattern 1 (monthly export):
#     YYYYMM_CYL07-{A2|MS}_{Result|RTime|Area}_airmoVOC {suffix}.Asc
#     e.g.: 202508_CYL07-A2_Result_airmoVOC C2-C6.Asc
#
#   Pattern 2 (date-range export):
#     YYYYMMDD_to_YYYYMMDD_CYL07-{A2|MS}_{Result|RTime|Area}.Asc
#     e.g.: 20250804_to_20250814_CYL07-A2_Result.Asc
#     Files in this format are assumed to match the user-specified compound_type.
#
# Both patterns can coexist in the same folder; all matching files are returned.
#
# Args:
#   folder     : Path to the folder containing CYL07 ASC files
#   asc_suffix : "C2-C6" or "C6-C12" — used to filter pattern 1 files only
#
# Returns:
#   A tibble with columns: filepath, filename, instrument, data_type
# ------------------------------------------------------------------------------

parse_cyl_filenames <- function(folder, asc_suffix) {

  # List all .Asc files in the folder (case-insensitive extension)
  all_files <- list.files(folder, pattern = "\\.Asc$", full.names = TRUE,
                          ignore.case = TRUE)

  if (length(all_files) == 0) {
    stop("No .Asc files found in folder: ", folder)
  }

  filenames <- basename(all_files)

  # ---------- Pattern 1: YYYYMM_CYL07-{inst}_{type}_airmoVOC {suffix}.Asc ----------
  pattern1 <- paste0(
    "^\\d{6}",                         # YYYYMM (not captured)
    "_(CYL07-(?:A2|MS))",               # group 1: instrument
    "_(Result|RTime|Area)",              # group 2: data type
    "_airmoVOC ",
    gsub("-", "\\\\-", asc_suffix), # compound suffix, hyphens escaped
    "\\.Asc$"
  )

  matches1 <- str_match(filenames, pattern1)
  valid1   <- !is.na(matches1[, 1])

  # ---------- Pattern 2: YYYYMMDD_to_YYYYMMDD_CYL07-{inst}_{type}.Asc ----------
  pattern2 <- paste0(
    "^\\d{8}_to_\\d{8}",   # date range prefix (not captured)
    "_(CYL07-(?:A2|MS))",      # group 1: instrument
    "_(Result|RTime|Area)",     # group 2: data type
    "\\.Asc$"
  )

  matches2 <- str_match(filenames, pattern2)
  valid2   <- !is.na(matches2[, 1])

  # ---------- Check that at least one file matched either pattern ----------
  if (sum(valid1) == 0 && sum(valid2) == 0) {
    stop(
      "No matching CYL07 ASC files found in folder: ", folder,
      "\nExpected one of:\n",
      "  Pattern 1: YYYYMM_CYL07-A2_Result_airmoVOC ", asc_suffix, ".Asc\n",
      "  Pattern 2: YYYYMMDD_to_YYYYMMDD_CYL07-A2_Result.Asc\n",
      "Check compound_type setting and filenames."
    )
  }

  # ---------- Build results tibble from whichever pattern(s) matched ----------
  results <- bind_rows(
    if (sum(valid1) > 0) tibble(
      filepath   = all_files[valid1],
      filename   = filenames[valid1],
      instrument = matches1[valid1, 2],
      data_type  = matches1[valid1, 3]
    ),
    if (sum(valid2) > 0) tibble(
      filepath   = all_files[valid2],
      filename   = filenames[valid2],
      instrument = matches2[valid2, 2],
      data_type  = matches2[valid2, 3]
    )
  )

  cat("Files matched - Pattern 1 (monthly):", sum(valid1),
      "| Pattern 2 (date-range):", sum(valid2), "\n")

  results
}
```

```{r helper_load_data_type}
# ------------------------------------------------------------------------------
# load_data_type()
#
# For a given data type ("Result", "RTime", or "Area"), reads all matching ASC
# files from the parsed file index, parses the datetime column, and combines
# them into a single long-format tibble with an instrument column.
#
# Args:
#   file_index  : Tibble produced by parse_cyl_filenames()
#   type        : One of "Result", "RTime", "Area"
#
# Returns:
#   A tibble with columns: instrument, datetime, Sampling date (raw), Volume,
#   and one column per compound (as character).
# ------------------------------------------------------------------------------

load_data_type <- function(file_index, type) {

  # Filter index to the requested data type
  subset_index <- file_index %>% filter(data_type == type)

  if (nrow(subset_index) == 0) {
    warning("No ", type, " files found in the file index.")
    return(NULL)
  }

  results <- list()

  for (i in seq_len(nrow(subset_index))) {

    row  <- subset_index[i, ]
    data <- read_asc_file(row$filepath)

    if (is.null(data)) {
      warning("Could not read file: ", row$filename)
      next
    }

    # Add instrument identifier as the first column
    data <- data %>%
      mutate(instrument = row$instrument, .before = 1)

    results[[i]] <- data
  }

  if (length(results) == 0) {
    warning("No ", type, " files could be read.")
    return(NULL)
  }

  # Combine all files for this data type
  bind_rows(results)
}
```

---

## Step 1: Discover ASC Files

```{r discover_files}
# Scan the CYL folder and parse metadata from filenames
file_index <- parse_cyl_filenames(cyl_folder, asc_suffix)

cat("Files found matching compound type '", asc_suffix, "':\n\n", sep = "")

file_index %>%
  select(filename, instrument, data_type) %>%
  arrange(instrument, data_type, filename) %>%
  print(n = Inf)

cat("\nInstruments detected:", paste(unique(file_index$instrument), collapse = ", "), "\n")
cat("Data types detected: ", paste(unique(file_index$data_type),  collapse = ", "), "\n\n")
```

---

## Step 2: Load Reference Compounds

```{r load_reference}
# Read the reference compound list from ID_cyl_peaks.csv.
# The file must contain a column named "Compound".

reference_raw <- read_csv(reference_file, show_col_types = FALSE)

if (!"Compound" %in% names(reference_raw)) {
  stop(
    "Reference file must contain a column named 'Compound'.\n",
    "Columns found: ", paste(names(reference_raw), collapse = ", ")
  )
}

reference_list <- reference_raw$Compound

cat("Reference compounds loaded (", length(reference_list), "):\n", sep = "")
cat(paste(reference_list, collapse = ", "), "\n\n")
```

---

## Step 3: Load ASC Data

```{r load_asc_data}
# Load all RTime and Result files from the file index.
# Each data type is combined across all months and both instruments.

cat("Loading RTime files...\n")
all_rtime <- load_data_type(file_index, "RTime")

cat("\nLoading Result files...\n")
all_result <- load_data_type(file_index, "Result")

# ---------- Validate ----------
if (is.null(all_rtime)) stop("No RTime data could be loaded. Check file names and folder.")
if (is.null(all_result)) stop("No Result data could be loaded. Check file names and folder.")

cat("\nRTime rows loaded: ", nrow(all_rtime),  "\n")
cat("Result rows loaded:", nrow(all_result), "\n\n")
```

---

## Step 4: RTime Comparison — Each Instrument vs Reference List

This section evaluates each instrument (A2 and MS) independently. For every
sampling row, it checks which expected compounds are present (RTime > 0),
which are absent, and which detections were not in the reference list.

```{r rtime_comparison}
# ---------- Identify compound columns ----------
# Exclude known metadata columns; everything else is a compound
metadata_cols <- c("instrument", "Sampling date", "source", "Volume")

compound_cols <- setdiff(names(all_rtime), metadata_cols)
cat("Compound columns found in RTime data:", length(compound_cols), "\n\n")

# ---------- Replace NA with 0 in compound columns ----------
all_rtime <- all_rtime %>%
  mutate(across(all_of(compound_cols), ~ replace_na(as.numeric(.x), 0)))

# ---------- Row-wise comparison against reference list ----------
# For each row (one sampling event per instrument):
#   - present_compounds      : compounds where RTime > 0
#   - absent_compounds       : compounds where RTime == 0
#   - expected_and_present   : reference compounds that were detected
#   - unexpected_and_present : detected compounds NOT in the reference list
#   - expected_but_absent    : reference compounds that were NOT detected

comparison_results <- all_rtime %>%
  select(instrument, sampling_date = `Sampling date`,
         all_of(compound_cols)) %>%
  rowwise() %>%
  mutate(
    analysis = list({
      row_data <- pick(all_of(compound_cols))

      present_compounds      <- names(row_data)[row_data > 0]
      absent_compounds       <- names(row_data)[row_data == 0]

      expected_and_present   <- intersect(present_compounds, reference_list)
      unexpected_and_present <- setdiff(present_compounds,   reference_list)
      expected_but_absent    <- intersect(absent_compounds,  reference_list)

      tibble(
        total_compounds_in_data      = length(compound_cols),
        compounds_with_values        = length(present_compounds),
        expected_compounds_present   = length(expected_and_present),
        unexpected_compounds_present = length(unexpected_and_present),
        expected_compounds_missing   = length(expected_but_absent),
        perfect_match = (length(unexpected_and_present) == 0) &
                        (length(expected_but_absent)    == 0),
        unexpected_list = ifelse(
          length(unexpected_and_present) > 0,
          paste(unexpected_and_present, collapse = "; "),
          NA_character_
        ),
        missing_list = ifelse(
          length(expected_but_absent) > 0,
          paste(expected_but_absent, collapse = "; "),
          NA_character_
        )
      )
    })
  ) %>%
  ungroup() %>%
  unnest(analysis)

cat("RTime comparison complete.\n\n")
```

---

## Step 5: RTime Summary

```{r rtime_summary}
cat("=== OVERALL SUMMARY ===\n\n")
cat("Total sampling rows analyzed:", nrow(comparison_results), "\n")
cat("Instruments analyzed:        ", paste(unique(comparison_results$instrument), collapse = ", "), "\n")
cat("Reference compounds:         ", length(reference_list), "\n")
cat("Compound columns in data:    ", length(compound_cols), "\n\n")

# ---------- Perfect match rate by instrument ----------
cat("=== PERFECT MATCH SUMMARY ===\n\n")
perfect_match_summary <- comparison_results %>%
  group_by(instrument) %>%
  summarise(
    total_samples   = n(),
    perfect_matches = sum(perfect_match),
    pct_perfect     = round(100 * perfect_matches / total_samples, 1),
    .groups = "drop"
  )
print(perfect_match_summary)
cat("\n")

# ---------- Average compound counts by instrument ----------
cat("=== COMPOUND PRESENCE SUMMARY ===\n\n")
compound_summary <- comparison_results %>%
  group_by(instrument) %>%
  summarise(
    avg_compounds_with_values = round(mean(compounds_with_values), 1),
    avg_expected_present      = round(mean(expected_compounds_present), 1),
    avg_unexpected_present    = round(mean(unexpected_compounds_present), 1),
    avg_expected_missing      = round(mean(expected_compounds_missing), 1),
    .groups = "drop"
  )
print(compound_summary)
cat("\n")

# ---------- Most common unexpected detections ----------
cat("=== MOST COMMON UNEXPECTED COMPOUNDS ===\n\n")
unexpected_compounds <- comparison_results %>%
  filter(!is.na(unexpected_list)) %>%
  separate_rows(unexpected_list, sep = "; ") %>%
  count(instrument, unexpected_list, sort = TRUE) %>%
  rename(compound = unexpected_list, occurrences = n)

if (nrow(unexpected_compounds) > 0) {
  print(unexpected_compounds)
} else {
  cat("No unexpected compounds found.\n")
}
cat("\n")

# ---------- Most common missing expected compounds ----------
cat("=== MOST COMMON MISSING COMPOUNDS ===\n\n")
missing_compounds <- comparison_results %>%
  filter(!is.na(missing_list)) %>%
  separate_rows(missing_list, sep = "; ") %>%
  count(instrument, missing_list, sort = TRUE) %>%
  rename(compound = missing_list, occurrences = n)

if (nrow(missing_compounds) > 0) {
  print(missing_compounds)
} else {
  cat("No missing expected compounds found.\n")
}
cat("\n")
```

---

## Step 6: Detailed RTime Results by Instrument

```{r rtime_detailed}
# Print a detailed breakdown for each instrument showing which sampling rows
# had mismatches against the reference list.

for (instr in sort(unique(comparison_results$instrument))) {

  cat("\n", strrep("=", 80), "\n", sep = "")
  cat("DETAILED RESULTS FOR:", instr, "\n")
  cat(strrep("=", 80), "\n\n", sep = "")

  instr_results <- comparison_results %>% filter(instrument == instr)

  cat("Total samples:       ", nrow(instr_results), "\n")
  cat("Perfect matches:     ", sum(instr_results$perfect_match), "\n")
  cat("Samples with issues: ", sum(!instr_results$perfect_match), "\n\n")

  problematic <- instr_results %>%
    filter(!perfect_match) %>%
    select(sampling_date,
           expected_compounds_present,
           unexpected_compounds_present,
           expected_compounds_missing,
           unexpected_list,
           missing_list)

  if (nrow(problematic) > 0) {
    cat("SAMPLES WITH MISALIGNMENTS:\n\n")
    print(problematic, n = Inf)
  } else {
    cat("All samples matched the reference perfectly.\n")
  }
}
```

---

## Step 7: Result Comparison — A2 vs MS

This section compares A2 and MS concentrations for shared reference compounds
on matching dates. Pairs where the symmetric percent difference exceeds the
threshold are flagged.

```{r result_comparison}
# ---------- Helper: round datetime to nearest day ----------
# A2 and MS timestamps may differ by ~1 hour; rounding to day allows matching.
round_to_day <- function(df) {
  df %>%
    mutate(
      date_raw = ymd_hms(`Sampling date`, truncated = 3),
      date_day = as_date(round_date(date_raw, unit = "day"))
    )
}

# ---------- Helper: pivot Result data to long format ----------
# Keeps only compounds that appear in the reference list.
pivot_result_long <- function(df, instrument_id) {

  meta_cols <- c("instrument", "Sampling date", "source", "Volume",
                 "date_raw", "date_day")
  comp_cols <- setdiff(names(df), meta_cols)

  # Restrict to reference compounds present in this file
  ref_cols_present <- intersect(comp_cols, reference_list)
  missing_ref      <- setdiff(reference_list, comp_cols)

  if (length(missing_ref) > 0) {
    cat("[", instrument_id, "] Reference compounds not in Result file: ",
        paste(missing_ref, collapse = ", "), "\n")
  }

  df %>%
    # Convert compound columns to numeric before pivoting
    mutate(across(all_of(ref_cols_present), as.numeric)) %>%
    select(instrument, date_day, all_of(ref_cols_present)) %>%
    pivot_longer(
      cols      = all_of(ref_cols_present),
      names_to  = "compound",
      values_to = "concentration"
    )
}

# ---------- Prepare A2 and MS Result data ----------
# Split all_result by instrument, then round dates and pivot to long format.

a2_result <- all_result %>% filter(instrument == "CYL07-A2")
ms_result <- all_result %>% filter(instrument == "CYL07-MS")

if (nrow(a2_result) == 0) warning("No A2 Result rows found.")
if (nrow(ms_result) == 0) warning("No MS Result rows found.")

a2_result_long <- a2_result %>% round_to_day() %>% pivot_result_long("CYL07-A2")
ms_result_long <- ms_result %>% round_to_day() %>% pivot_result_long("CYL07-MS")

# ---------- Join on date + compound ----------
result_joined <- inner_join(
  a2_result_long %>% rename(conc_a2 = concentration) %>% select(-instrument),
  ms_result_long %>% rename(conc_ms = concentration) %>% select(-instrument),
  by = c("date_day", "compound")
)

if (nrow(result_joined) == 0) {
  cat("WARNING: No matching date/compound pairs found between A2 and MS Result files.\n")
} else {

  # ---------- Calculate symmetric percent difference ----------
  # Formula: abs(A2 - MS) / mean(A2, MS) * 100
  # Skipped (NA) when both values are zero (no meaningful comparison)
  result_joined <- result_joined %>%
    mutate(
      mean_conc = (conc_a2 + conc_ms) / 2,
      pct_diff  = ifelse(
        mean_conc == 0, NA_real_,
        abs(conc_a2 - conc_ms) / mean_conc * 100
      ),
      flagged   = !is.na(pct_diff) & pct_diff > pct_diff_threshold
    )

  # ---------- Print summary ----------
  cat("=== RESULT COMPARISON: A2 vs MS ===\n\n")
  cat("Threshold:                  ", pct_diff_threshold, "% difference\n")
  cat("Matched date/compound pairs:", nrow(result_joined), "\n")
  cat("Pairs skipped (both = 0):  ", sum(is.na(result_joined$pct_diff)), "\n")
  cat("Flagged pairs (>", pct_diff_threshold, "% diff):",
      sum(result_joined$flagged, na.rm = TRUE), "\n\n")

  flagged_results <- result_joined %>%
    filter(flagged) %>%
    arrange(desc(pct_diff))

  if (nrow(flagged_results) > 0) {
    cat("--- FLAGGED PAIRS (sorted by % difference) ---\n\n")
    print(
      flagged_results %>%
        mutate(across(c(conc_a2, conc_ms, mean_conc, pct_diff), ~ round(.x, 4))),
      n = Inf
    )
    cat("\n")

    cat("--- MOST FREQUENTLY FLAGGED COMPOUNDS ---\n\n")
    flag_freq <- flagged_results %>%
      count(compound, sort = TRUE) %>%
      rename(times_flagged = n) %>%
      mutate(pct_of_dates = round(
        100 * times_flagged / n_distinct(result_joined$date_day), 1
      ))
    print(flag_freq)
    cat("\n")
  } else {
    cat("No pairs exceeded the", pct_diff_threshold, "% threshold.\n\n")
  }
}
```

---

## Step 8: Continuity Check

This section checks for unexpected gaps between sampling dates and large
concentration jumps between consecutive measurements. Each instrument is
evaluated separately.

```{r continuity_check}
# ---------- Combine long Result data from both instruments ----------
all_result_long <- bind_rows(a2_result_long, ms_result_long) %>%
  arrange(instrument, compound, date_day)

# ---------- Gap detection ----------
# Flag consecutive measurements separated by more than gap_threshold_days.
gap_flags <- all_result_long %>%
  group_by(instrument, compound) %>%
  arrange(date_day, .by_group = TRUE) %>%
  mutate(
    prev_date   = lag(date_day),
    days_since  = as.numeric(date_day - prev_date),
    gap_flagged = !is.na(days_since) & days_since > gap_threshold_days
  ) %>%
  ungroup() %>%
  filter(gap_flagged) %>%
  select(instrument, compound,
         gap_start = prev_date, gap_end = date_day, days_gap = days_since)

cat("=== CONTINUITY CHECK: DATE GAPS ===\n\n")
cat("Gap threshold:", gap_threshold_days, "days\n\n")

if (nrow(gap_flags) > 0) {
  cat("Gaps detected:\n\n")
  print(gap_flags %>% arrange(instrument, compound, gap_start), n = Inf)
} else {
  cat("No gaps exceeding", gap_threshold_days, "days detected.\n")
}
cat("\n")

# ---------- Concentration jump detection ----------
# Flag consecutive measurements where the symmetric % change exceeds the
# pct_diff_threshold (reuses the same threshold as the A2 vs MS comparison).
jump_flags <- all_result_long %>%
  group_by(instrument, compound) %>%
  arrange(date_day, .by_group = TRUE) %>%
  mutate(
    prev_conc   = lag(concentration),
    prev_date   = lag(date_day),
    mean_consec = (concentration + prev_conc) / 2,
    pct_change  = ifelse(
      is.na(prev_conc) | mean_consec == 0, NA_real_,
      abs(concentration - prev_conc) / mean_consec * 100
    ),
    jump_flagged = !is.na(pct_change) & pct_change > pct_diff_threshold
  ) %>%
  ungroup() %>%
  filter(jump_flagged) %>%
  select(instrument, compound,
         date_from = prev_date, conc_from = prev_conc,
         date_to   = date_day,  conc_to   = concentration,
         pct_change)

cat("=== CONTINUITY CHECK: CONCENTRATION JUMPS ===\n\n")
cat("Jump threshold:", pct_diff_threshold,
    "% change between consecutive measurements\n\n")

if (nrow(jump_flags) > 0) {
  cat("Concentration jumps detected:\n\n")
  print(
    jump_flags %>%
      mutate(across(c(conc_from, conc_to, pct_change), ~ round(.x, 4))) %>%
      arrange(instrument, compound, date_from),
    n = Inf
  )
} else {
  cat("No concentration jumps exceeding", pct_diff_threshold, "% detected.\n")
}
cat("\n")
```

---

## Step 9: Export RTime Comparison Report

```{r export_rtime}
# Export the RTime comparison results to CSV.
# One row per sampling event per instrument, with columns summarizing
# presence/absence of expected and unexpected compounds.

final_comparison <- comparison_results %>%
  select(
    instrument,
    sampling_date,
    expected_compounds_present,
    unexpected_compounds_present,
    expected_compounds_missing,
    perfect_match,
    unexpected_list,
    missing_list
  )

rtime_export_path <- file.path(output_folder, "rtime_comparison_report.csv")
write_csv(final_comparison, rtime_export_path)

cat("=== EXPORT COMPLETE ===\n\n")
cat("RTime comparison report saved to:\n", rtime_export_path, "\n")
cat("Rows exported:", nrow(final_comparison), "\n")
```
