---
title: "C6C12 RT Adjustment File Formatter"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    code_folding: show
    theme: cosmo
---

## Overview

This script processes C6C12 retention time (RT) adjustment Excel files and
formats them for use with the MethodManager Python automation script. It is
designed to be run on a folder of input files, one per calibration date.

**What this script does:**

1. Reads input Excel files (one sheet named "valid") containing compound names
   and observed R.Time values from a GC calibration run
2. Joins observed R.Time values to a hardcoded reference table of all 92 C6C12
   compounds and their baseline RT windows (OldMin, OldMax, OldMid)
3. Calculates the RT shift (adjustment) as: observed R.Time - OldMid
4. Applies the shift to calculate new RT windows (New Rt MIN, New Rt MAX)
5. Sets New Rt MIN and New Rt MAX to 0 for compounds absent from the input file
6. Writes output Excel files formatted for the MethodManager Python script

**Python script requirements for output files:**

- Row 3: Column headers (Python reads headers from this row)
- Rows 4-7: Empty (spacer rows expected by Python)
- Rows 8-99: All 92 C6C12 compounds, starting with ACETALDEHYDE
- Column named "Name": compound names (Python identifies compounds here)
- Column named "New Rt MIN": RT minimum values (Python finds by "RT"+"MIN")
- Column named "New Rt MAX": RT maximum values (Python finds by "RT"+"MAX")
- Values of 0 for MIN/MAX tell Python to skip that compound in MethodManager

**Dependencies (files needed to run this script):**

- Input folder containing one or more Excel files matching the pattern:
  `C6C12_YYYY_MM_DD.xlsx` (e.g., C6C12_2024_03_15.xlsx)
- Each input file must have a sheet named "valid" (configurable below) with:
  - A column named "Name" containing compound names matching the reference table
  - A column named "R.Time" containing observed retention times (in seconds)
  - Column headers in row 3 (rows 1-2 are skipped on read)
- No external reference files are required; the 92-compound table is built in

---

## User Configuration

**Edit only this section before running.**

```{r setup, echo=TRUE, warning=FALSE, message=FALSE}

# ------------------------------------------------------------------------------
# LOAD REQUIRED LIBRARIES
# ------------------------------------------------------------------------------
# tidyverse: provides dplyr (data manipulation), stringr (string ops), readr
# readxl:   reads .xlsx input files
# writexl:  writes .xlsx output files (use writexl, NOT openxlsx, for Python
#           compatibility — openxlsx adds drawing XML that breaks openpyxl)
# ------------------------------------------------------------------------------

library(tidyverse)
library(readxl)
library(writexl)

# ==============================================================================
# USER CONFIGURATION — edit these parameters before running
# ==============================================================================

# Full path to the folder containing your C6C12_YYYY_MM_DD.xlsx input files.
# Use forward slashes (/) even on Windows, or escaped backslashes (\\).
# Example: "C:/Users/yourname/Documents/RT Adjustment C6C12"
input_dir <- "YOUR/INPUT/FOLDER/PATH/HERE"

# Output directory for processed files.
# NULL means output files will be saved to the same folder as the input files.
# To save elsewhere, replace NULL with a quoted path string.
output_dir <- NULL

# Regex pattern used to identify valid input files in the input folder.
# Default matches filenames like: C6C12_2024_03_15.xlsx
# Only change this if your file naming convention differs.
file_pattern <- "^C6C12_\\d{4}_\\d{2}_\\d{2}\\.xlsx$"

# Name of the sheet in each input Excel file that contains compound data.
# Default is "valid". Change if your sheet is named differently.
input_sheet <- "valid"

# Column name in the input sheet containing compound names.
# These names must exactly match the reference table built into this script.
input_name_col <- "Name"

# Column name in the input sheet containing observed retention times (seconds).
input_rtime_col <- "R.Time"

# ==============================================================================
```

---

## Reference Compound Table

This table is hardcoded and defines all 92 C6C12 compounds along with their
baseline RT window boundaries. It serves as the "master list" that every input
file is compared against.

- **OldMin / OldMax**: baseline RT window edges (seconds)
- **DRt**: window width (OldMax - OldMin)
- **OldMid**: baseline window midpoint; used to calculate the RT shift

Compounds present in the input file will have their windows shifted by:
`adjustment = round(observed R.Time - OldMid)`

Compounds absent from the input file will have New Rt MIN = 0 and New Rt MAX = 0,
which tells the Python script to skip them in MethodManager.

```{r reference_table, echo=TRUE, warning=FALSE, message=FALSE}

# Build the reference table for all 92 C6C12 compounds.
# Column order: Name, OldMin, OldMax, DRt (window width), OldMid (midpoint).
# Do not reorder or rename columns — downstream calculations depend on these names.

ref_compounds <- tibble::tribble(
  ~Name,                ~OldMin, ~OldMax, ~DRt, ~OldMid,
  "ACETALDEHYDE",       14,      22,      8,    18,
  "CHLOROETHANE",       22,      30,      8,    26,
  "CHLOROMETHANE",      30,      38,      8,    34,
  "N-BUTANE",           47,      57,      10,   52,
  "ACROLEIN",           86,      96,      10,   91,
  "ACETONE",            96,      106,     10,   101,
  "PROPANAL",           111,     113,     2,    112,
  "N-PENTANE",          113,     123,     10,   118,
  "DICHLOROMETHANE",    185,     195,     10,   190,
  "2,2-DIME-BUTANE",    195,     215,     20,   205,
  "2-METHYLPROPENAL",   210,     220,     10,   215,
  "1,1-DICHLOROETHANE", 250,     260,     10,   255,
  "MTBE",               257,     267,     10,   262,
  "VINYLACETATE",       265,     275,     10,   270,
  "2-ME-PENTANE",       275,     285,     10,   280,
  "T-1-2-DICL-ETHENE",  285,     295,     10,   290,
  "MEK",                304,     314,     10,   309,
  "C-1,2-DICL-ETHENE",  353,     363,     10,   358,
  "N-HEXANE",           366,     376,     10,   371,
  "CHLOROFORME",        375,     385,     10,   380,
  "THF",                429,     439,     10,   434,
  "CROTONALDEHYDE",     451,     461,     10,   456,
  "ME-CYCLOPENTANE",    467,     477,     10,   472,
  "1-2DICHLOROETHANE",  473,     483,     10,   478,
  "111-TriCL-ETHANE",   503,     513,     10,   508,
  "BENZENE",            562,     582,     20,   572,
  "CCI4",               582,     602,     20,   592,
  "CYCLOHEXANE",        602,     622,     20,   612,
  "2-ME-HEXANE",        640,     650,     10,   645,
  "3-ME-HEXANE",        681,     691,     10,   686,
  "1-2-DICL-PROPANE",   699,     709,     10,   704,
  "PENTANAL",           709,     712,     3,    710.5,
  "BROMO-DICL-METHAN",  726,     737,     11,   731.5,
  "TRICHLOROETHENE",    736,     745,     9,    740.5,
  "224-TME-PENTANE",    740,     750,     10,   745,
  "1-4-DIOXANE",        745,     750,     5,    747.5,
  "ME-METHACRYLATE",    776,     786,     10,   781,
  "N-HEPTANE",          786,     798,     12,   792,
  "T-1-3-DICL-PROPEN",  876,     888,     12,   882,
  "ME-CYCLOHEXANE",     884,     894,     10,   889,
  "MIBK",               886,     890,     4,    888,
  "C-1-3-DICL-PROPEN",  954,     970,     16,   962,
  "112-TriCL-ETHANE",   976,     992,     16,   984,
  "234-TME-PENTANE",    990,     1010,    20,   1000,
  "TOLUENE",            1020,    1030,    10,   1025,
  "2-ME-HEPTANE",       1050,    1064,    14,   1057,
  "MBK",                1065,    1075,    10,   1070,
  "DIBR-CL-METHANE",    1078,    1084,    6,    1081,
  "3-ME-HEPTANE",       1078,    1090,    12,   1084,
  "HEXANAL",            1089,    1099,    10,   1094,
  "1-2DIBROMOETHANE",   1109,    1119,    10,   1114,
  "N-OCTANE",           1165,    1177,    12,   1171,
  "TETRACHLORORETHENE", 1176,    1186,    10,   1181,
  "FURFURAL",           1238,    1248,    10,   1243,
  "CHLOROBENZENE",      1263,    1283,    20,   1273,
  "ETHYLBENZENE",       1315,    1335,    20,   1325,
  "M&P-XYLENES",        1344,    1364,    20,   1354,
  "BROMOFORM",          1355,    1367,    12,   1361,
  "STYRENE",            1391,    1403,    12,   1397,
  "O-XYLENE",           1404,    1416,    12,   1410,
  "1122TetraCL-ETHAN",  1416,    1422,    6,    1419,
  "N-NONANE",           1435,    1455,    20,   1445,
  "I-PROPYLBENZENE",    1478,    1498,    20,   1488,
  "BENZALDEHYDE",       1510,    1520,    10,   1515,
  "a-PINENE",           1530,    1542,    12,   1536,
  "2-CHLORO-TOLUENE",   1544,    1554,    10,   1549,
  "N-PROPYLBENZENE",    1550,    1560,    10,   1555,
  "M-ETHYLTOLUENE",     1566,    1574,    8,    1570,
  "P-ETHYLTOLUENE",     1572,    1580,    8,    1576,
  "135-TMB",            1580,    1594,    14,   1587,
  "O-ETHYLTOLUENE",     1605,    1619,    14,   1612,
  "b-PINENE",           1618,    1632,    14,   1625,
  "124-TMB",            1637,    1647,    10,   1642,
  "BENZYLCHLORIDE",     1659,    1663,    4,    1661,
  "1-3-DICL-BENZENE",   1659,    1663,    4,    1661,
  "N-DECANE",           1657,    1667,    10,   1662,
  "1-4-DICL-BENZENE",   1667,    1677,    10,   1672,
  "3-CARENE",           1685,    1695,    10,   1690,
  "123-TMB",            1696,    1708,    12,   1702,
  "1-2-DICL-BENZENE",   1716,    1726,    10,   1721,
  "LIMONENE",           1720,    1730,    10,   1725,
  "M-DIETHYLBENZENE",   1745,    1755,    10,   1750,
  "P-DIETHYLBENZENE",   1759,    1769,    10,   1764,
  "M-TOLUALDEHYDE",     1769,    1779,    10,   1774,
  "N-UNDECANE",         1838,    1858,    20,   1848,
  "124-TriCL-BENZENE",  2006,    2020,    14,   2013,
  "NAPHTHALENE",        2030,    2040,    10,   2035,
  "N-DODECANE",         2038,    2058,    20,   2048,
  "HexCL-1-3BUTADIEN",  2106,    2126,    20,   2116,
  "N-TRIDECANE",        2174,    2194,    20,   2184,
  "2-ME-NAPHTHALENE",   2254,    2274,    20,   2264,
  "1-ME-NAPHTHALENE",   2299,    2319,    20,   2309
)

# Quick sanity check — should print 92 compounds, first = ACETALDEHYDE,
# last = 1-ME-NAPHTHALENE. If these differ, the reference table has been altered.
cat("Reference table contains", nrow(ref_compounds), "compounds\n")
cat("First compound:", ref_compounds$Name[1], "\n")
cat("Last compound: ", ref_compounds$Name[nrow(ref_compounds)], "\n")
```

---

## File Discovery

```{r file_discovery, echo=TRUE, warning=FALSE, message=FALSE}

# ------------------------------------------------------------------------------
# RESOLVE OUTPUT DIRECTORY
# If output_dir was left as NULL in User Configuration, save outputs alongside
# the input files. Otherwise, create the specified output directory if needed.
# ------------------------------------------------------------------------------

if (is.null(output_dir)) {
  output_dir <- input_dir
} else if (!dir.exists(output_dir)) {
  dir.create(output_dir, recursive = TRUE)
  cat("Created output directory:", output_dir, "\n")
}

# ------------------------------------------------------------------------------
# CHECK INPUT DIRECTORY EXISTS
# Stops immediately with a clear message if the path is wrong, rather than
# producing a cryptic downstream error.
# ------------------------------------------------------------------------------

if (!dir.exists(input_dir)) {
  stop(
    "Input directory not found: ", input_dir,
    "\nPlease check the input_dir path in the User Configuration section."
  )
}

# ------------------------------------------------------------------------------
# FIND MATCHING INPUT FILES
# list.files() searches the input folder for filenames matching file_pattern.
# full.names = TRUE returns complete file paths for use in read_excel() later.
# ignore.case = TRUE makes the match case-insensitive (e.g., .XLSX = .xlsx).
# ------------------------------------------------------------------------------

all_files <- list.files(
  path        = input_dir,
  pattern     = file_pattern,
  full.names  = TRUE,
  ignore.case = TRUE
)

# Extract just filenames for display (not full paths)
file_names <- basename(all_files)

cat(paste0(strrep("=", 62), "\n"))
cat("FILE DISCOVERY\n")
cat(paste0(strrep("=", 62), "\n"))
cat("Input directory:", input_dir, "\n")
cat("File pattern:   ", file_pattern, "\n")
cat("Files found:    ", length(all_files), "\n\n")

# Stop if no files match — better to fail early with a clear message than to
# silently produce no output
if (length(all_files) == 0) {
  stop(
    "No files found matching pattern '", file_pattern, "' in:\n", input_dir,
    "\nCheck that your files are named C6C12_YYYY_MM_DD.xlsx ",
    "and that input_dir is correct."
  )
}

cat("Files to process:\n")
for (i in seq_along(file_names)) {
  cat(sprintf("  %2d. %s\n", i, file_names[i]))
}
cat("\n")
```

---

## Processing Functions

Three functions handle the core logic. They are defined here and called in the
next section. Separating functions from the processing loop makes it easier to
debug individual steps.

```{r functions, echo=TRUE, warning=FALSE, message=FALSE}

# ==============================================================================
# FUNCTION 1: validate_compounds()
#
# Checks that every compound name in the input file exactly matches a name in
# the reference table. Mismatches will cause a left_join() to silently drop
# rows, producing incorrect output — so we catch them here first.
#
# Arguments:
#   input_compounds  character vector of compound names read from the input file
#   ref_compounds    the 92-row reference tibble defined above
#
# Returns a list with:
#   $valid       TRUE if all input names are in the reference table
#   $input_count number of unique compound names in the input file
#   $not_in_ref  character vector of input names that have no reference match
#                (empty if $valid is TRUE)
# ==============================================================================

validate_compounds <- function(input_compounds, ref_compounds) {

  input_names <- unique(input_compounds)
  ref_names   <- ref_compounds$Name

  # setdiff(A, B) returns elements in A that are not in B
  not_in_ref <- setdiff(input_names, ref_names)

  list(
    valid       = length(not_in_ref) == 0,
    input_count = length(input_names),
    not_in_ref  = not_in_ref
  )
}


# ==============================================================================
# FUNCTION 2: process_file()
#
# Reads one input Excel file, validates compound names, calculates RT
# adjustments, and returns a formatted 92-row output tibble.
#
# Arguments:
#   file_path       full path to the input .xlsx file
#   ref_compounds   the 92-row reference tibble
#   input_sheet     name of the sheet to read (default: "valid")
#   input_name_col  column name containing compound names (default: "Name")
#   input_rtime_col column name containing observed RT values (default: "R.Time")
#
# Returns a list with:
#   $file_name       basename of the input file
#   $success         TRUE if processing completed without errors
#   $error           error message string if $success is FALSE, else NULL
#   $compounds_found number of compounds matched between input and reference
#   $compounds_total total compounds in reference table (always 92)
#   $output_data     formatted 92-row tibble ready to write (if $success TRUE)
# ==============================================================================

process_file <- function(file_path,
                         ref_compounds,
                         input_sheet,
                         input_name_col,
                         input_rtime_col) {

  file_name <- basename(file_path)

  cat("\n", strrep("=", 62), "\n", sep = "")
  cat("Processing:", file_name, "\n")
  cat(strrep("=", 62), "\n")

  # Initialise a result object. Default is failure; overwritten on success.
  result <- list(
    file_name       = file_name,
    success         = FALSE,
    error           = NULL,
    compounds_found = 0,
    compounds_total = nrow(ref_compounds),
    output_data     = NULL
  )

  # Wrap everything in tryCatch so one bad file doesn't abort the whole batch.
  # On error the message is stored in result$error and we move on.
  tryCatch({

    # --------------------------------------------------------------------------
    # STEP 1: Read the input sheet
    # skip = 2 skips rows 1-2 so row 3 becomes the header row, matching the
    # Vistachrom export format expected by this workflow.
    # --------------------------------------------------------------------------
    cat("Reading sheet '", input_sheet, "' from input file...\n", sep = "")

    input_data <- read_excel(
      path  = file_path,
      sheet = input_sheet,
      skip  = 2   # Rows 1-2 skipped; row 3 treated as column headers
    )

    # Confirm required columns are present before attempting to use them
    if (!input_name_col %in% names(input_data)) {
      stop("Column '", input_name_col, "' not found. ",
           "Available columns: ", paste(names(input_data), collapse = ", "))
    }
    if (!input_rtime_col %in% names(input_data)) {
      stop("Column '", input_rtime_col, "' not found. ",
           "Available columns: ", paste(names(input_data), collapse = ", "))
    }

    # --------------------------------------------------------------------------
    # STEP 2: Extract and clean compound data
    # Select only the two columns we need, drop any rows missing Name or R.Time
    # (blank rows at the bottom of the sheet), and trim whitespace from names.
    # --------------------------------------------------------------------------
    compound_data <- input_data %>%
      select(
        Name   = all_of(input_name_col),
        R.Time = all_of(input_rtime_col)
      ) %>%
      filter(!is.na(Name), !is.na(R.Time)) %>%
      mutate(Name = str_trim(Name))   # Remove leading/trailing spaces

    cat("Compounds found in input file:", nrow(compound_data), "\n")

    # --------------------------------------------------------------------------
    # STEP 3: Validate compound names
    # Every name in the input must exactly match the reference table. Even one
    # mismatch (e.g. extra space, different capitalisation) will cause that
    # compound to be treated as absent (zeroed out) — so we stop and report it.
    # --------------------------------------------------------------------------
    validation <- validate_compounds(compound_data$Name, ref_compounds)

    if (!validation$valid) {
      cat("\n*** VALIDATION ERROR: Unrecognised compound names ***\n")
      cat("The following names in the input file do not match the reference table:\n")
      for (nm in validation$not_in_ref) {
        cat("  - '", nm, "'\n", sep = "")
      }
      cat("\nTip: Check for extra spaces, capitalisation differences, or typos.\n")
      cat("Reference table names are all-caps with specific punctuation.\n\n")
      stop("Compound name validation failed. See messages above.")
    }

    cat("All compound names validated successfully\n")

    # --------------------------------------------------------------------------
    # STEP 4: Calculate RT adjustments
    #
    # Logic:
    #   - Left-join observed R.Time onto the full 92-compound reference table.
    #     Compounds absent from the input file will have NA in R.Time after join.
    #   - For matched compounds: adjustment = round(observed R.Time - OldMid)
    #                            New Rt MIN  = OldMin + adjustment
    #                            New Rt MAX  = OldMax + adjustment
    #   - For unmatched compounds: all new values set to 0 (signals Python to skip)
    #
    # Column layout (A-M) is fixed — Python identifies columns by header string:
    #   A  Name             — Python reads compound names here
    #   B  OldMin           — baseline window lower bound (reference, not read by Python)
    #   C  OldMax           — baseline window upper bound (reference, not read by Python)
    #   D  DRt              — window width (reference, not read by Python)
    #   E  OldMid           — baseline midpoint (reference, not read by Python)
    #   F  New Adjustment   — shift applied (seconds)
    #   G  R.Time           — observed RT from input (0 if compound absent)
    #   H  RtMID - R.Time   — difference between new midpoint and observed RT
    #   I  Adjustment (s)   — duplicate of F for display consistency
    #   J  New Rt MIN       — Python reads this (searches for "RT" + "MIN")
    #   K  New Rt MAX       — Python reads this (searches for "RT" + "MAX")
    #   L  New Rt MID       — new window midpoint
    #   M  New RtMID - R.Time — verification: should be near 0 for matched compounds
    # --------------------------------------------------------------------------
    cat("Calculating RT adjustments...\n")

    output_data <- ref_compounds %>%

      # Left join: keeps all 92 reference rows; adds R.Time where matched
      left_join(compound_data, by = "Name") %>%

      mutate(

        # Shift = difference between observed RT and baseline midpoint, rounded
        # to nearest integer second. NA (absent compound) -> 0.
        `New Adjustment` = if_else(
          !is.na(R.Time),
          round(R.Time - OldMid),
          0
        ),

        # New window lower bound: shift baseline min by the adjustment
        `New Rt MIN` = if_else(
          !is.na(R.Time),
          OldMin + `New Adjustment`,
          0   # 0 tells Python to skip this compound
        ),

        # New window upper bound: shift baseline max by the adjustment
        `New Rt MAX` = if_else(
          !is.na(R.Time),
          OldMax + `New Adjustment`,
          0   # 0 tells Python to skip this compound
        ),

        # New midpoint: average of new min and max
        `New Rt MID` = if_else(
          !is.na(R.Time),
          (`New Rt MIN` + `New Rt MAX`) / 2,
          0
        ),

        # Replace NA R.Time (absent compounds) with 0 for clean output
        R.Time = if_else(is.na(R.Time), 0, R.Time),

        # Difference between new midpoint and observed RT — should be ~0
        # for all matched compounds (non-zero indicates rounding)
        `RtMID - R.Time` = `New Rt MID` - R.Time,

        # Duplicate adjustment column — kept for display consistency with
        # the original file format
        `Adjustment (s)` = `New Adjustment`,

        `New RtMID - R.Time` = `New Rt MID` - R.Time

      ) %>%

      # Reorder columns to match the expected A-M layout (see header above)
      select(
        Name,                  # A
        OldMin,                # B
        OldMax,                # C
        DRt,                   # D
        OldMid,                # E
        `New Adjustment`,      # F
        R.Time,                # G
        `RtMID - R.Time`,      # H
        `Adjustment (s)`,      # I
        `New Rt MIN`,          # J — Python reads this
        `New Rt MAX`,          # K — Python reads this
        `New Rt MID`,          # L
        `New RtMID - R.Time`   # M
      )

    # Count how many compounds had an adjustment applied (non-zero shift)
    compounds_found <- sum(output_data$`New Adjustment` != 0)
    cat("Compounds with RT adjustments:", compounds_found, "\n")
    cat("Compounds set to 0 (absent from input):",
        nrow(output_data) - compounds_found, "\n")

    # --------------------------------------------------------------------------
    # STEP 5: Verify output structure
    # Python will fail silently if row count or first compound are wrong,
    # so we check here and stop with a clear message if something is off.
    # --------------------------------------------------------------------------
    cat("\nVerifying output structure...\n")
    cat("  Total rows:", nrow(output_data), "(expected: 92)\n")
    cat("  First compound:", output_data$Name[1], "(expected: ACETALDEHYDE)\n")
    cat("  Last compound: ", output_data$Name[nrow(output_data)], "\n")

    if (nrow(output_data) != 92) {
      stop("Output has ", nrow(output_data), " rows but must have exactly 92. ",
           "Check the reference table for duplicate or missing entries.")
    }

    if (output_data$Name[1] != "ACETALDEHYDE") {
      stop("First compound is '", output_data$Name[1],
           "' but must be ACETALDEHYDE. ",
           "Do not reorder the reference table.")
    }

    cat("Output structure validated\n")

    # All checks passed — update result to success
    result$success         <- TRUE
    result$compounds_found <- compounds_found
    result$output_data     <- output_data

  }, error = function(e) {
    # Store the error message and return; the loop will continue to the next file
    result$error <<- conditionMessage(e)
    cat("\n*** ERROR:", conditionMessage(e), "\n")
  })

  return(result)
}


# ==============================================================================
# FUNCTION 3: write_output_file()
#
# Writes a processed 92-row output tibble to an Excel file with the exact row
# placement required by the Python MethodManager script:
#   Row 1-2: Empty (NA)
#   Row 3:   Column headers (written as data, not as R's default header row)
#   Row 4-7: Empty (NA)
#   Row 8-99: The 92 compound data rows
#
# This structure is achieved by building a single flat data frame where the
# header row is inserted as a data row, then writing with col_names = FALSE
# to suppress R's automatic header. All values are coerced to character first
# to prevent type conflicts when binding rows of mixed numeric/character data.
#
# Why writexl, not openxlsx?
#   openxlsx adds a drawing XML element that causes errors when the file is
#   later read by Python's openpyxl library. writexl produces a clean file.
#
# Arguments:
#   output_data     the 92-row tibble returned by process_file()
#   input_file_path full path to the original input file (used for naming)
#   output_dir      folder to write the output file into
#
# Returns the full path of the written output file.
# ==============================================================================

write_output_file <- function(output_data, input_file_path, output_dir) {

  # Derive output filename by appending "_all" to the input filename stem
  # e.g. C6C12_2024_03_15.xlsx -> C6C12_2024_03_15_all.xlsx
  input_stem  <- tools::file_path_sans_ext(basename(input_file_path))
  output_name <- paste0(input_stem, "_all.xlsx")
  output_path <- file.path(output_dir, output_name)

  cat("\nWriting output file...\n")
  cat("  Path:", output_path, "\n")

  # --------------------------------------------------------------------------
  # Convert all columns to character before binding rows of different types.
  # numeric + character bind will coerce silently and may corrupt values.
  # --------------------------------------------------------------------------
  output_char <- output_data %>%
    mutate(across(everything(), as.character))

  n_cols    <- ncol(output_char)
  col_names <- names(output_char)   # The actual column name strings (A-M labels)

  # Helper to create a block of empty (NA) rows with the same column structure
  make_empty_rows <- function(n) {
    as.data.frame(
      matrix(NA_character_, nrow = n, ncol = n_cols),
      stringsAsFactors = FALSE
    ) %>%
      setNames(col_names)
  }

  # Build a one-row data frame containing the column names as data values.
  # This becomes row 3 in the output file, where Python looks for headers.
  header_row <- as.data.frame(
    matrix(col_names, nrow = 1, ncol = n_cols),
    stringsAsFactors = FALSE
  ) %>%
    setNames(col_names)

  # Stack all sections into one flat data frame:
  #   Rows 1-2:  empty
  #   Row 3:     column names (as data)
  #   Rows 4-7:  empty
  #   Rows 8-99: compound data (92 rows)
  final_data <- bind_rows(
    make_empty_rows(2),   # Rows 1-2
    header_row,           # Row 3  — Python reads headers here
    make_empty_rows(4),   # Rows 4-7
    output_char           # Rows 8-99
  )

  # Write to Excel.
  # col_names = FALSE prevents writexl from adding its own header row above
  # our manually placed header in row 3.
  # Sheet is named "C6C12" to match convention; Python does not check sheet name.
  write_xlsx(
    x         = list(C6C12 = final_data),
    path      = output_path,
    col_names = FALSE
  )

  cat("  File written successfully\n")
  cat("  Structure: headers in row 3, compound data in rows 8-99\n")

  return(output_path)
}
```

---

## Process All Files

This chunk calls the two processing functions for every file discovered above.
Results are collected in a summary table for display in the next section.

```{r process_files, echo=TRUE, warning=FALSE, message=FALSE}

# Initialise an empty summary table. One row will be added per input file.
processing_summary <- tibble(
  file_name       = character(),
  success         = logical(),
  compounds_found = integer(),
  compounds_total = integer(),
  error           = character()
)

# --------------------------------------------------------------------------
# MAIN PROCESSING LOOP
# Iterates over every file path found in the file discovery step.
# process_file() handles its own errors via tryCatch, so a failure on one
# file will not abort the remaining files.
# --------------------------------------------------------------------------

for (file_path in all_files) {

  # STEP 1: Read, validate, and compute adjustments
  result <- process_file(
    file_path       = file_path,
    ref_compounds   = ref_compounds,
    input_sheet     = input_sheet,
    input_name_col  = input_name_col,
    input_rtime_col = input_rtime_col
  )

  # STEP 2: Write output file only if processing succeeded
  if (result$success) {
    write_output_file(
      output_data     = result$output_data,
      input_file_path = file_path,
      output_dir      = output_dir
    )
  }

  # STEP 3: Append this file's result to the summary table
  processing_summary <- processing_summary %>%
    bind_rows(
      tibble(
        file_name       = result$file_name,
        success         = result$success,
        compounds_found = as.integer(result$compounds_found),
        compounds_total = as.integer(result$compounds_total),
        error           = ifelse(is.null(result$error), "", result$error)
      )
    )
}

cat("\n")
```

---

## Processing Summary

```{r summary, echo=TRUE, warning=FALSE, message=FALSE}

cat("\n")
cat(paste0(strrep("=", 70), "\n"))
cat("PROCESSING SUMMARY\n")
cat(paste0(strrep("=", 70), "\n\n"))

cat("Files processed:", nrow(processing_summary), "\n")
cat("Successful:     ", sum(processing_summary$success), "\n")
cat("Failed:         ", sum(!processing_summary$success), "\n\n")

cat("File Details:\n")
cat(strrep("-", 70), "\n")

for (i in seq_len(nrow(processing_summary))) {
  row    <- processing_summary[i, ]
  symbol <- if_else(row$success, "OK", "FAIL")

  cat(sprintf("[%s] %s\n", symbol, row$file_name))

  if (row$success) {
    cat(sprintf("     Compounds adjusted: %d / %d\n",
                row$compounds_found, row$compounds_total))
    cat(sprintf("     Compounds zeroed:   %d / %d\n",
                row$compounds_total - row$compounds_found, row$compounds_total))
  } else {
    cat(sprintf("     ERROR: %s\n", row$error))
  }
}

cat(strrep("-", 70), "\n")

# Aggregate statistics across all successful files
if (sum(processing_summary$success) > 0) {
  cat("\nAggregate (successful files only):\n")
  cat("  Total compound slots processed:",
      sum(processing_summary$compounds_total[processing_summary$success]), "\n")
  cat("  Total with adjustments applied:",
      sum(processing_summary$compounds_found[processing_summary$success]), "\n")
  cat("  Total zeroed out:              ",
      sum(
        (processing_summary$compounds_total - processing_summary$compounds_found
        )[processing_summary$success]
      ), "\n")
}

cat("\n")
cat(paste0(strrep("=", 70), "\n"))
cat("Output directory:", output_dir, "\n")
cat(paste0(strrep("=", 70), "\n"))
```

---

## Python Compatibility Notes

This section documents the output file structure so future users understand
why the script is formatted the way it is.

### Output file naming
Input file `C6C12_YYYY_MM_DD.xlsx` → output file `C6C12_YYYY_MM_DD_all.xlsx`,
saved to the same folder unless `output_dir` is changed.

### Row layout

| Rows   | Contents                                      |
|--------|-----------------------------------------------|
| 1-2    | Empty (reserved)                              |
| 3      | Column headers — Python reads headers here    |
| 4-7    | Empty (spacer rows expected by Python)        |
| 8-99   | All 92 C6C12 compounds in reference order     |

### Column layout (A-M)

| Column | Header              | Read by Python? | Notes                              |
|--------|---------------------|-----------------|------------------------------------|
| A      | Name                | Yes             | Compound identification            |
| B      | OldMin              | No              | Baseline window lower bound        |
| C      | OldMax              | No              | Baseline window upper bound        |
| D      | DRt                 | No              | Window width                       |
| E      | OldMid              | No              | Baseline midpoint                  |
| F      | New Adjustment      | No              | Shift in seconds                   |
| G      | R.Time              | No              | Observed RT from input             |
| H      | RtMID - R.Time      | No              | Difference (should be ~0)          |
| I      | Adjustment (s)      | No              | Duplicate of F                     |
| J      | New Rt MIN          | Yes             | Python finds via "RT" + "MIN"      |
| K      | New Rt MAX          | Yes             | Python finds via "RT" + "MAX"      |
| L      | New Rt MID          | No              | New window midpoint                |
| M      | New RtMID - R.Time  | No              | Verification column                |

### Why 0 for missing compounds?

Compounds absent from the input file receive `New Rt MIN = 0` and
`New Rt MAX = 0`. The Python script reads these zeros as a signal to skip
that compound when updating MethodManager — it will not overwrite existing
RT windows for compounds that were not observed in this calibration run.

### First compound check

The Python script verifies that row 8 contains `ACETALDEHYDE` before
proceeding. Do not reorder the reference table in this script.

### Package choice: writexl vs openxlsx

`writexl` is used intentionally. `openxlsx` adds a `xl/drawings/drawing1.xml`
element to the Excel file that causes errors when Python's `openpyxl` library
reads the file. `writexl` produces a minimal, clean `.xlsx` that openpyxl
handles without issue.

---

## Session Info

```{r session_info, echo=TRUE}
sessionInfo()
```
