---
title: "GC MDL Exceedance Analysis"
output: html_document
---

# Dependencies

**Upstream script:** `Combine_processed_data_w0_Claude.Rmd`

**Input files** (read from `output_dir` = `data_dir/output/`):

- `ZERO60-2_Result_combined.csv` — C2C6 zero check results
- `ZERO-MS_Result_combined.csv` — C6C12 zero check results

**MDL reference files** (read from `mdl_c2c6` and `mdl_c6c12` paths); units specified by `mdl_units` parameter ("ppt" or "ppb")

**Output:** None — diagnostic only, results printed to console.

---

## User-Specified Parameters

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(readr)
library(lubridate)

# ===== USER PARAMETERS =====
base_dir  <- ""   # Root directory containing ASC files (use / not \)
subfolder <- ""   # Optional seasonal subfolder (e.g. "Winter2425"); leave "" if files are directly in base_dir

mdl_c2c6  <- ""   # Full path to C2C6 MDL file  (use / not \)
mdl_c6c12 <- ""   # Full path to C6C12 MDL file (use / not \)

compound_types <- c("C6C12")  # Options: "C2C6", "C6C12", or c("C2C6", "C6C12")
mdl_units      <- "ppb"       # Units of MDL file: "ppt" or "ppb"
# ===========================

# Derived paths
data_dir   <- file.path(base_dir, subfolder)
output_dir <- file.path(data_dir, "output")
```

## Process Overview

This analysis compares measured compound values against Method Detection Limits (MDL) to identify any exceedances. The workflow:

1. Load result files from `output_dir` and MDL reference files
2. Exclude non-compound columns from comparison
3. Compare each compound measurement against its MDL
4. Report only instances where measured values exceed MDL

------------------------------------------------------------------------

## C2C6 Analysis

```{r c2c6-analysis, message=FALSE, warning=FALSE}
if ("C2C6" %in% compound_types) {
  
  # Define file path
  c2c6_result_file <- file.path(output_dir, "ZERO60-2_Result_combined.csv")
  
  # Check if files exist
  if (!file.exists(c2c6_result_file)) {
    cat("ERROR: C2C6 result file not found at:", c2c6_result_file, "\n")
    c2c6_exceedances <- NULL
  } else if (!file.exists(mdl_c2c6)) {
    cat("ERROR: C2C6 MDL file not found at:", mdl_c2c6, "\n")
    c2c6_exceedances <- NULL
  } else {
    
    # Read data files
    c2c6_results <- read_csv(c2c6_result_file, show_col_types = FALSE)
    c2c6_mdl     <- read_csv(mdl_c2c6, show_col_types = FALSE)
    
    # Convert MDL units to ppb if needed (results are in ppb)
    if (mdl_units == "ppt") {
      c2c6_mdl <- c2c6_mdl %>% mutate(across(where(is.numeric), ~ .x / 1000))
      cat("MDL values converted from ppt to ppb (÷1000)\n")
    }
    
    # Columns to exclude from comparison
    exclude_cols_c2c6 <- c("Sampling date", "source", "Volume", "NMTHC-EQC4")
    
    # Get compound columns (all except excluded ones)
    compound_cols_c2c6 <- setdiff(names(c2c6_results), exclude_cols_c2c6)
    
    # Initialize exceedances dataframe
    c2c6_exceedances <- tibble()
    
    for (i in 1:nrow(c2c6_results)) {
      sampling_date <- c2c6_results$`Sampling date`[i]
      
      for (compound in compound_cols_c2c6) {
        if (!compound %in% names(c2c6_mdl)) next
        
        measured_value <- c2c6_results[[compound]][i]
        mdl_value      <- c2c6_mdl[[compound]][1]  # MDL file has single row
        
        if (!is.na(measured_value) && !is.na(mdl_value) && measured_value > mdl_value) {
          c2c6_exceedances <- c2c6_exceedances %>%
            bind_rows(tibble(
              Compound_Type  = "C2C6",
              Sampling_Date  = sampling_date,
              Compound       = compound,
              Measured_Value = measured_value,
              MDL_Value      = mdl_value,
              Difference     = measured_value - mdl_value
            ))
        }
      }
    }
    
    cat("\n### C2C6 Results\n\n")
    cat("Total measurements analyzed:", nrow(c2c6_results) * length(compound_cols_c2c6), "\n")
    cat("Number of MDL exceedances:", nrow(c2c6_exceedances), "\n\n")
    
    if (nrow(c2c6_exceedances) > 0) {
      cat("MDL Exceedances Found:\n\n")
      print(c2c6_exceedances, n = Inf)
    } else {
      cat("✓ No MDL exceedances detected for C2C6 compounds.\n")
    }
  }
} else {
  cat("C2C6 analysis skipped (not in compound_types parameter).\n")
  c2c6_exceedances <- NULL
}
```

------------------------------------------------------------------------

## C6C12 Analysis

```{r c6c12-analysis, message=FALSE, warning=FALSE}
if ("C6C12" %in% compound_types) {
  
  # Define file path
  c6c12_result_file <- file.path(output_dir, "ZERO-MS_Result_combined.csv")
  
  # Check if files exist
  if (!file.exists(c6c12_result_file)) {
    cat("ERROR: C6C12 result file not found at:", c6c12_result_file, "\n")
    c6c12_exceedances <- NULL
  } else if (!file.exists(mdl_c6c12)) {
    cat("ERROR: C6C12 MDL file not found at:", mdl_c6c12, "\n")
    c6c12_exceedances <- NULL
  } else {
    
    # Read data files
    c6c12_results <- read_csv(c6c12_result_file, show_col_types = FALSE)
    c6c12_mdl     <- read_csv(mdl_c6c12, show_col_types = FALSE)
    
    # Convert MDL units to ppb if needed (results are in ppb)
    if (mdl_units == "ppt") {
      c6c12_mdl <- c6c12_mdl %>% mutate(across(where(is.numeric), ~ .x / 1000))
      cat("MDL values converted from ppt to ppb (÷1000)\n")
    }
    
    # Columns to exclude from comparison
    exclude_cols_c6c12 <- c("Sampling date", "source", "Volume", "NMTHC-EQC6")
    
    # Get compound columns (all except excluded ones)
    compound_cols_c6c12 <- setdiff(names(c6c12_results), exclude_cols_c6c12)
    
    # Initialize exceedances dataframe
    c6c12_exceedances <- tibble()
    
    for (i in 1:nrow(c6c12_results)) {
      sampling_date <- c6c12_results$`Sampling date`[i]
      
      for (compound in compound_cols_c6c12) {
        if (!compound %in% names(c6c12_mdl)) next
        
        measured_value <- c6c12_results[[compound]][i]
        mdl_value      <- c6c12_mdl[[compound]][1]  # MDL file has single row
        
        if (!is.na(measured_value) && !is.na(mdl_value) && measured_value > mdl_value) {
          c6c12_exceedances <- c6c12_exceedances %>%
            bind_rows(tibble(
              Compound_Type  = "C6C12",
              Sampling_Date  = sampling_date,
              Compound       = compound,
              Measured_Value = measured_value,
              MDL_Value      = mdl_value,
              Difference     = measured_value - mdl_value
            ))
        }
      }
    }
    
    cat("\n### C6C12 Results\n\n")
    cat("Total measurements analyzed:", nrow(c6c12_results) * length(compound_cols_c6c12), "\n")
    cat("Number of MDL exceedances:", nrow(c6c12_exceedances), "\n\n")
    
    if (nrow(c6c12_exceedances) > 0) {
      cat("MDL Exceedances Found:\n\n")
      print(c6c12_exceedances, n = Inf)
    } else {
      cat("✓ No MDL exceedances detected for C6C12 compounds.\n")
    }
  }
} else {
  cat("C6C12 analysis skipped (not in compound_types parameter).\n")
  c6c12_exceedances <- NULL
}
```

------------------------------------------------------------------------

## Combined Summary

```{r combined-summary, message=FALSE, warning=FALSE}
# Combine all exceedances
all_exceedances <- bind_rows(c2c6_exceedances, c6c12_exceedances)

if (!is.null(all_exceedances) && nrow(all_exceedances) > 0) {
  cat("\n### All MDL Exceedances Combined\n\n")
  cat("Total exceedances across all compound types:", nrow(all_exceedances), "\n\n")
  
  exceedance_summary <- all_exceedances %>%
    group_by(Compound_Type) %>%
    summarise(
      Number_of_Exceedances = n(),
      Unique_Compounds      = n_distinct(Compound),
      Unique_Dates          = n_distinct(Sampling_Date)
    )
  
  print(exceedance_summary)
  
  cat("\n### Detailed Exceedances Table\n\n")
  print(all_exceedances, n = Inf)
  
} else {
  cat("\n✓ No MDL exceedances detected across all analyzed compound types.\n")
}
```
