---
title: "MDL Replacement for Ambient Data"
output: html_document
---

## Purpose

This script replaces zero values in ambient VOC measurement data with 1/2 the Method Detection Limit (MDL). Compounds are filtered to those present in both the ambient data and MDL file (intersection-based). The workflow:

1. Load ambient and MDL data
2. Filter to compounds present in both files
3. Evaluate zero percentages for reference and quality assessment
4. Validate column matching between datasets
5. Replace zeros with 1/2 MDL values (converting units if needed)
6. Export processed data with summary statistics

---

# Dependencies

**Upstream scripts:** `Combine_processed_data_w0_Claude.Rmd` → `exp_data_comb_Claude.Rmd`

**Input files** (read from `output_dir` = `data_dir/output/`):

- `AMB60_{compound_type}_{subfolder}_combined.csv` — merged ambient data

**MDL reference file:** full path specified by `mdl_file`

**Output files** (written to `output_dir`):

- `AMB60_{compound_type}_{subfolder}_MDLrep_34c.csv` — ambient data with zeros replaced by 1/2 MDL

---

## User Parameters

```{r user-parameters, message=FALSE, warning=FALSE}
library(tidyverse)
library(lubridate)

# ===== USER PARAMETERS =====
base_dir      <- ""       # Root directory containing ASC files
subfolder     <- ""       # Optional seasonal subfolder (e.g. "Winter2425"); leave "" if files are directly in base_dir

compound_type <- "C6C12"  # "C2C6" or "C6C12"

mdl_file      <- ""       # Full path to MDL file
mdl_units     <- "ppb"    # Units of MDL file: "ppt" or "ppb"

# Zero percentage threshold — for reporting/flagging only, does not affect filtering
zero_cutoff_percent <- 99.99

# Non-compound columns to exclude from processing
exclude_cols <- c("Sampling date", "Volume", "NMTHC-EQC6", "source")

# Additional compounds to manually exclude by name (e.g. compounds known to be unreliable)
# Set to c() to skip manual exclusions
manual_exclude <- c()
# ===========================

# Derived paths
data_dir   <- file.path(base_dir, subfolder)
output_dir <- file.path(data_dir, "output")

# Derived filenames
amb_filename    <- paste0("AMB60_", compound_type, "_", subfolder, "_combined.csv")
output_filename <- paste0("AMB60_", compound_type, "_", subfolder, "_MDLrep_34c.csv")

amb_path    <- file.path(output_dir, amb_filename)
output_path <- file.path(output_dir, output_filename)
```

## Load Data

```{r load-data, message=FALSE, warning=FALSE}
if (!file.exists(amb_path)) stop("Ambient file not found: ", amb_path)
if (!file.exists(mdl_file)) stop("MDL file not found: ", mdl_file)

cat("Loading ambient file:", amb_filename, "\n")
amb_data <- read_csv(amb_path, show_col_types = FALSE)

cat("Loading MDL file:", basename(mdl_file), "\n")
mdl_raw <- read_csv(mdl_file, show_col_types = FALSE)

cat("\nAmbient data dimensions:", nrow(amb_data), "rows x", ncol(amb_data), "columns\n")
cat("MDL data dimensions:", nrow(mdl_raw), "rows x", ncol(mdl_raw), "columns\n")
```

## Evaluate Zero Percentages and Filter by MDL

Zero percentages are calculated for reference. Compounds are retained only if present in both the ambient data and the MDL file.

```{r zero-evaluation, message=FALSE, warning=FALSE}
# Identify compound columns (exclude metadata and manual exclusions)
compound_cols <- setdiff(names(amb_data), c(exclude_cols, manual_exclude))

cat("Total compound columns:", length(compound_cols), "\n")

# Calculate zero percentages for each compound
zero_stats <- amb_data %>%
  select(all_of(compound_cols)) %>%
  summarise(across(everything(),
                   ~sum(. == 0, na.rm = TRUE) / n() * 100)) %>%
  pivot_longer(everything(),
               names_to = "Compound",
               values_to = "Zero_Percent")

cat("\n=== Zero Percentage Summary ===\n")
zero_stats %>% arrange(desc(Zero_Percent)) %>% print(n = Inf)

# Filter to compounds present in both ambient data and MDL file
mdl_compounds      <- names(mdl_raw)
compounds_to_keep  <- intersect(compound_cols, mdl_compounds)
compounds_to_remove <- setdiff(compound_cols, mdl_compounds)

cat("\n=== Filtering Results ===\n")
cat("Compounds in ambient data:", length(compound_cols), "\n")
cat("Compounds in MDL file:", length(mdl_compounds), "\n")
cat("Compounds retained (in both):", length(compounds_to_keep), "\n")
cat("Compounds removed (not in MDL):", length(compounds_to_remove), "\n")
if (length(compounds_to_remove) > 0) {
  cat("Removed compounds:\n")
  cat(paste("  -", compounds_to_remove, collapse = "\n"), "\n")
}

# Flag retained compounds with high zero percentages
zero_stats_kept <- zero_stats %>% filter(Compound %in% compounds_to_keep)

cat("\n=== Zero Percentage Stats for Retained Compounds ===\n")
cat("Compounds with >", zero_cutoff_percent, "% zeros:",
    sum(zero_stats_kept$Zero_Percent > zero_cutoff_percent), "\n")
if (sum(zero_stats_kept$Zero_Percent > zero_cutoff_percent) > 0) {
  zero_stats_kept %>%
    filter(Zero_Percent > zero_cutoff_percent) %>%
    arrange(desc(Zero_Percent)) %>%
    print(n = Inf)
}

# Filter ambient data to retained compounds
amb_filtered <- amb_data %>%
  select(`Sampling date`, all_of(compounds_to_keep))
```

## Process MDL Data

```{r process-mdl, message=FALSE, warning=FALSE}
# Convert to long format
mdl_processed_all <- mdl_raw %>%
  pivot_longer(everything(),
               names_to = "Compound",
               values_to = "MDL_raw") %>%
  mutate(
    MDL_ppb      = if (mdl_units == "ppt") MDL_raw / 1000 else MDL_raw,
    Half_MDL_ppb = MDL_ppb / 2
  )

cat("=== MDL Units ===\n")
cat("Input units:", mdl_units, "\n")
if (mdl_units == "ppt") cat("MDL values converted from ppt to ppb (÷1000)\n")

cat("\nTotal compounds in MDL file:", nrow(mdl_processed_all), "\n")

# Report compounds in MDL but not in filtered ambient data
mdl_only_compounds <- setdiff(mdl_processed_all$Compound, compounds_to_keep)
if (length(mdl_only_compounds) > 0) {
  cat("\nNOTE: Compounds in MDL but NOT in filtered ambient data (",
      length(mdl_only_compounds), "):\n")
  cat(paste("  -", mdl_only_compounds, collapse = "\n"), "\n")
}

# Filter to matching compounds only
mdl_processed <- mdl_processed_all %>%
  filter(Compound %in% compounds_to_keep)

cat("\n=== Processed MDL Values (Matching Compounds Only) ===\n")
print(mdl_processed, n = Inf)
```

## Column Matching Validation

```{r column-validation, message=FALSE, warning=FALSE}
amb_compounds <- compounds_to_keep
mdl_compounds <- mdl_processed$Compound

in_amb_not_mdl <- setdiff(amb_compounds, mdl_compounds)
in_mdl_not_amb <- setdiff(mdl_compounds, amb_compounds)

cat("=== Column Matching Validation ===\n")
cat("Compounds in filtered ambient data:", length(amb_compounds), "\n")
cat("Compounds in MDL data (filtered):", length(mdl_compounds), "\n")

if (length(in_amb_not_mdl) > 0) {
  cat("\n*** CRITICAL MISMATCH DETECTED ***\n")
  cat("Compounds in ambient but NOT in MDL (", length(in_amb_not_mdl), "):\n")
  cat(paste("  -", in_amb_not_mdl, collapse = "\n"), "\n")
  stop("*** PROCESSING STOPPED: Ambient compounds missing from MDL file ***\n")
} else {
  cat("\n✓ All ambient compound columns have corresponding MDL values\n")
}
```

## MDL Replacement

Replace all zero values with 1/2 MDL for each compound.

```{r mdl-replacement, message=FALSE, warning=FALSE}
amb_replaced <- amb_filtered

replacement_stats <- tibble(
  Compound          = character(),
  Zeros_Before      = integer(),
  Zeros_After       = integer(),
  Replacements_Made = integer(),
  Half_MDL_Value    = numeric()
)

for (compound in mdl_processed$Compound) {
  half_mdl <- mdl_processed %>%
    filter(Compound == !!compound) %>%
    pull(Half_MDL_ppb)
  
  zeros_before <- sum(amb_replaced[[compound]] == 0, na.rm = TRUE)
  
  amb_replaced[[compound]][amb_replaced[[compound]] == 0] <- half_mdl
  
  zeros_after <- sum(amb_replaced[[compound]] == 0, na.rm = TRUE)
  
  replacement_stats <- replacement_stats %>%
    bind_rows(tibble(
      Compound          = compound,
      Zeros_Before      = zeros_before,
      Zeros_After       = zeros_after,
      Replacements_Made = zeros_before - zeros_after,
      Half_MDL_Value    = half_mdl
    ))
}

cat("=== MDL Replacement Summary ===\n")
cat("Total replacements made:", sum(replacement_stats$Replacements_Made), "\n\n")
print(replacement_stats, n = Inf)
```

## Before/After Statistics

```{r before-after-stats, message=FALSE, warning=FALSE}
stats_before <- amb_filtered %>%
  select(-`Sampling date`) %>%
  summarise(across(everything(),
                   list(
                     min    = ~min(., na.rm = TRUE),
                     mean   = ~mean(., na.rm = TRUE),
                     median = ~median(., na.rm = TRUE),
                     max    = ~max(., na.rm = TRUE),
                     zeros  = ~sum(. == 0, na.rm = TRUE)
                   ))) %>%
  pivot_longer(everything(),
               names_to = c("Compound", "Statistic"),
               names_sep = "_",
               values_to = "Before")

stats_after <- amb_replaced %>%
  select(-`Sampling date`) %>%
  summarise(across(everything(),
                   list(
                     min    = ~min(., na.rm = TRUE),
                     mean   = ~mean(., na.rm = TRUE),
                     median = ~median(., na.rm = TRUE),
                     max    = ~max(., na.rm = TRUE),
                     zeros  = ~sum(. == 0, na.rm = TRUE)
                   ))) %>%
  pivot_longer(everything(),
               names_to = c("Compound", "Statistic"),
               names_sep = "_",
               values_to = "After")

comparison_stats <- stats_before %>%
  left_join(stats_after, by = c("Compound", "Statistic")) %>%
  mutate(Change = After - Before) %>%
  arrange(Compound, Statistic)

cat("=== Before/After Statistics Comparison ===\n")
print(comparison_stats, n = 50)

cat("\n=== Key Changes ===\n")
cat("Minimum values changed:",
    sum(comparison_stats %>% filter(Statistic == "min") %>% pull(Change) != 0),
    "compounds\n")
cat("All zeros eliminated:",
    all(comparison_stats %>% filter(Statistic == "zeros", After == 0) %>% nrow() ==
        comparison_stats %>% filter(Statistic == "zeros") %>% nrow()), "\n")
```

## Export Processed Data

```{r export, message=FALSE, warning=FALSE}
write_csv(amb_replaced, output_path)

cat("=== Export Complete ===\n")
cat("Output file:", output_filename, "\n")
cat("Dimensions:", nrow(amb_replaced), "rows x", ncol(amb_replaced), "columns\n")
cat("Columns: Sampling date +", ncol(amb_replaced) - 1, "compounds\n")
```

## Summary Report

```{r summary-report, message=FALSE, warning=FALSE}
cat("\n", paste(rep("=", 60), collapse = ""), "\n")
cat("                   PROCESSING SUMMARY\n")
cat(paste(rep("=", 60), collapse = ""), "\n\n")

cat("Subfolder:", subfolder, "\n")
cat("Compound type:", compound_type, "\n")
cat("Input file:", amb_filename, "\n")
cat("Output file:", output_filename, "\n\n")

cat("--- Compound Filtering ---\n")
cat("Original compound count:", length(compound_cols), "\n")
cat("Compounds in MDL file:", length(names(mdl_raw)), "\n")
cat("Compounds removed (not in MDL):", length(compounds_to_remove), "\n")
cat("Compounds retained (in both files):", length(compounds_to_keep), "\n")
cat("Note: Zero cutoff (", zero_cutoff_percent, "%) used for flagging only\n\n")

cat("--- MDL Replacement ---\n")
cat("MDL units (input):", mdl_units, "\n")
cat("Total zero values replaced:", sum(replacement_stats$Replacements_Made), "\n")
cat("Compounds with replacements:", sum(replacement_stats$Replacements_Made > 0), "\n")
cat("MDL source file:", basename(mdl_file), "\n\n")

cat("--- Output ---\n")
cat("Final dataset dimensions:", nrow(amb_replaced), "rows x", ncol(amb_replaced), "columns\n")
cat("Date range:", format(min(amb_replaced$`Sampling date`)), "to",
    format(max(amb_replaced$`Sampling date`)), "\n")

cat("\n", paste(rep("=", 60), collapse = ""), "\n")
```

## Final Compound Statistics

```{r final-statistics, message=FALSE, warning=FALSE}
final_stats <- amb_replaced %>%
  select(-`Sampling date`) %>%
  summarise(across(everything(),
                   list(
                     min    = ~min(., na.rm = TRUE),
                     q25    = ~quantile(., 0.25, na.rm = TRUE),
                     median = ~median(., na.rm = TRUE),
                     mean   = ~mean(., na.rm = TRUE),
                     q75    = ~quantile(., 0.75, na.rm = TRUE),
                     max    = ~max(., na.rm = TRUE),
                     sd     = ~sd(., na.rm = TRUE),
                     n_obs  = ~sum(!is.na(.))
                   ))) %>%
  pivot_longer(everything(),
               names_to = c("Compound", "Statistic"),
               names_sep = "_",
               values_to = "Value") %>%
  pivot_wider(names_from = Statistic, values_from = Value) %>%
  arrange(Compound)

cat("=== Final Compound Statistics (After MDL Replacement) ===\n")
cat("Statistics for", nrow(final_stats), "compounds\n\n")
print(final_stats, n = Inf)

cat("\n=== Distribution Summary ===\n")
cat("Compounds with mean > median (right-skewed):",
    sum(final_stats$mean > final_stats$median, na.rm = TRUE), "\n")
cat("Compounds with mean < median (left-skewed):",
    sum(final_stats$mean < final_stats$median, na.rm = TRUE), "\n")
cat("Average coefficient of variation (SD/Mean):",
    round(mean(final_stats$sd / final_stats$mean, na.rm = TRUE), 2), "\n")
```
